Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-11-04 draft_lm1.tex
changed title. fixes issue #1
started rewriting the introduction, and added afew extra references to the bibliography

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Estimating the Initial Susceptible Fraction in Dengue Dynamics}
}
% Insert Author names, affiliations and corresponding author email.
\\
Flavio Code\c{c}o Coelho$^{1\ast}$, 
Luiz Max de Carvalho$^{2}$, 
\\
\bf{1} Escola de Matem\'atica Aplicada , Funda\c{c}\~ao Getulio Vargas (FGV), 
Rio de Janeiro -- RJ, Brazil.
\\
\bf{2} Programa de Computa\c{c}\~ao Cient\'ifica (PROCC), Funda\c{c}\~ao Oswaldo Cruz, Rio de Janeiro -- RJ, Brazil.
\\
$\ast$ E-mail: fccoelho@fgv.br
\end{flushleft}

@@

Dengue is a disease which is caused by 4 different types of viruses 
which co-circulate between the vector and human populations. immunity is 
considered permanent after the exposure of humans to each of the viruses but 
cross immunity is considered very limited[ref].
Thus, the proportions of viral types co-circulating at any point in time is 
strongly dependent on previous incidence of the disease, which determines the 
number of susceptibles for each viral type at any point in time.

Dengue transmission is conditioned on temperature, due to its effects on the 
vector reproduction.
In places with sufficient temperature variations dengue is a summer disease 
mainly.

Determining the viral types in circulation during an epidemic is 
expensive, as it requires virus isolation and reverse-transcriptase-PCR. 
Therefore only a very limited number of samples are examined on any epidemic. 
Besides, mild cases are rarely detected let alone tested for dengue.

Dengue in Brazil is a disease which is well in its way to 
endemicity but not yet stable, therefore co-circulation patterns have not yet 
reached a steady state as has been demonstrated for certain countries in 
south-east asia which have had the disease much longer.

In the particular case of Dengue fever in Brazil, with four circulating 
serotypes, it can be very challenging to parameterize a simulation starting at 
any time other than when the date the disease was reintroduced in the country 
in the late 1980s and even so, without virological survey data from time to 
time, to inform on the mixing of viruses in circulation, it is virtually 
impossible to predict, by simulation alone, the imunological state of the 
population at any point in the future.

For SIR-like models, the relevant initial values are the number of infectious 
and susceptible  individuals. 
The number of infectious individuals are less challenging to determine, because 
the get very small during the winters and can be derived from incidence data  
with is monitored continuously.
The real challenge is to determine the number of susceptibles to 
a particular strain right at the beginning of the dengue season.

Multiple factors contribute to the dificulty in estimating the number of 
susceptibles in the population, first and foremost is the fact that dengue
confers permanent immunity within serotypes but with only short-lived 
cross-immunity. 
Secondly, there are a larger number of asymptomatic cases in 
any epidemic, which nevertheless acquire immunity. It is estimated that for 
every case reported, 10-20 are not seen by health authorities (personnal 
communication).
Thirdly, under-reporting is a serious issue in Brazil. 
Duarte and Fran\c{c}a\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, 
meaning that aproximately 37\% of the suspected dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways which 
are not easy to determine.

>>>>

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Estimating the Attack ratio of Dengue Epidemics without serotype 
information}
}
% Insert Author names, affiliations and corresponding author email.
\\
Flavio Code\c{c}o Coelho$^{1\ast}$, 
Luiz Max de Carvalho$^{2}$, 
\\
\bf{1} Escola de Matem\'atica Aplicada , Funda\c{c}\~ao Getulio Vargas (FGV), 
Rio de Janeiro -- RJ, Brazil.
\\
\bf{2} Programa de Computa\c{c}\~ao Cient\'ifica (PROCC), Funda\c{c}\~ao Oswaldo Cruz, Rio de Janeiro -- RJ, Brazil.
\\
$\ast$ E-mail: fccoelho@fgv.br
\end{flushleft}

@@

Dengue is a disease which is caused by 4 different types of viruses 
which co-circulate between the vector and human populations. Immunity is 
considered permanent after the exposure of humans to each of the viruses and 
cross immunity is considered very limited[ref].
Thus, the proportions of viral types co-circulating at any point in time is 
strongly dependent on previous incidence of the disease, which determines the 
number of susceptibles for each viral type at any point in time.

Dengue transmission is modulated by environmental conditions, of which 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence\cite{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient temperature variation dengue is predominantly a 
summer disease.

The long term dynamics of dengue is also modulated by the alternation of virus 
types in circulation Demographics also plays a role in replenishing the 
population of susceptibles.

The Attack ratio of disease is a mesure of mobidity defined as the number of new
cases divided byt the population at risk. For Dengue outbreaks and epidemics 
this can be difficult to calculated due to the lack of information about the 
population at risk. The population at risk in the case of dengue, is the number 
of susceptibles to the circulating virus type(s) befor a given epidemic.

The main source of information about Dengue dynamics are the incidence 
timeseries, which represent all the cases of Dengue reported by health 
professionals with no discrimination of virus types. Only a very small sample 
of the cases are ever examined to determined the virus type.

In order to calculate the attack ratio, we need to determine the number of 
susceptibles to the circulating virus types right before the epidemic. Without 
regular virological surveys, it is virtually impossible to determine the 
population at risk.

The attack ratio is also influenced by the reproductive number of the disease 
\cite{bacaer_final_2009, katriel_attack_2012}. With Dengue, since transmission 
is affected by environmental conditions as discussed above, the reproductive 
number is a function of time.

%%%% Revised up to here

Secondly, there are a larger number of asymptomatic cases in 
any epidemic, which nevertheless acquire immunity. It is estimated that for 
every case reported, 10-20 are not seen by health authorities (personnal 
communication).
Thirdly, under-reporting is a serious issue in Brazil. 
Duarte and Fran\c{c}a\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, 
meaning that aproximately 37\% of the suspected dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways which 
are not easy to determine.
